---
title: " Introduction to Statistical Methods for Data Science"
output: html_notebook
---

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Ctrl+Shift+Enter*.

### Install required dependencies

```{r}
#install.packages("matlib")
#install.packages("rsample")
install.packages("dplyr")

```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

## Import Necessary Library

```{r}
library(matlib)
library(ggplot2)
library(rsample)
library(patchwork)
library(gridExtra)
library(grid)
```

## Importing data from csv

```{r}
set.seed(123)
df <- read.csv(file="~/assignment/stat/dataset.csv", header = TRUE)
any(is.na(df))
head(df)
```

## Splitting the data
```{r}
X <- as.matrix(df[, c("x1", "x3", "x4", "x5")])
Y <- as.matrix(df[,"x2"])
time <- seq(1, by = 1, length.out = nrow(df))
```

```{r}
head(time)
```

```{r}
X.ts<-ts(X,start = c(min(time),max(time)),frequency =1)
Y.ts<-ts(Y,start = c(min(time),max(time)),frequency =1)
```

```{r}
plot(X.ts, main = "Time series plot of X Signal", xlab = "Time", ylab = "Input signal", col = "steelblue")
plot(Y.ts, main = "Time series plot of Y Signal", xlab = "Time", ylab = "Output signal", col = "steelblue")
```

```{r}
plot_ts <- function(data, group_size = 100, n_groups = 3) {
  
  if(!"Index" %in% names(data)) {
    data$Index <- 1:nrow(data)
  }
  
  variables <- setdiff(names(data)[sapply(data, is.numeric)], "Index")
  
  for (var_name in variables) {
    
    # Group plots
    plot_list <- list()
    for (i in 1:n_groups) {
      start_idx <- (i-1) * group_size + 1
      end_idx <- min(start_idx + group_size - 1, nrow(data))
      subset_df <- data[start_idx:end_idx, ]
      
      p <- ggplot(subset_df, aes(x = Index, y = .data[[var_name]])) +
        geom_line(color = "darkorange") +
        labs(title = paste("Group", i), x = "Index", y = if(var_name == "x2") "Y" else var_name) +
        theme_minimal()
      
      plot_list[[i]] <- p
    }
    
    grid.arrange(grobs = plot_list, ncol = 1, top = paste("Groups -", if(var_name == "x2") "Y" else var_name))
  }
}
plot_ts(df)
```

```{r}
density_of_X=density(X)
plot(density_of_X,main = "Density plot of input signal X")
```

### Creating a Histogram of X signal

```{r}

```

```{r}
hist(X,freq = FALSE,main = "Density", breaks = 30)
```

#### Combining Histogram of X signal with density plot

```{r}
hist(X,freq = FALSE,main = "Density")
lines(density_of_X,lwd=2,col="red")
rug(jitter(X))
```

#### Histogram and density plot of individual input signal X and output signal y

```{r}
# Creating a density plot of input signal X1 
density_of_X1 <- density(X[,"x1"])

# Add 'breaks = 30' to control the number of bins
hist(X[,"x1"], freq = FALSE, breaks = 30, 
     main = "Histogram and Density plot of x1", xlab = "x1 Signal")

# Overlay density curve
lines(density_of_X1, lwd = 2, col = "red")

# Add the data-points with noise in the X-axis
rug(jitter(X[,"x1"]))

```

```{r}
density_of_X3=density(X[,"x3"])
hist(X[,"x3"],freq = FALSE,main = "Histogram and Density plot of x3",xlab = "x3 Signal")
lines(density_of_X3,lwd=2,col="red")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x3"]))
```

```{r}
#Creating a density plot of input signal X4
density_of_X4=density(X[,"x4"])
hist(X[,"x4"],freq = FALSE,main = "Histogram and Density plot of x4",xlab = "x4 Signal")
lines(density_of_X4,lwd=2,col="red")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x4"]))
```

```{r}
#Creating a density plot of input signal X5
density_of_X5=density(X[,"x5"])
hist(X[,"x5"],freq = FALSE,main = "Histogram and density plot of x5",xlab = "x5 Signal")
lines(density_of_X5,lwd=2,col="red")
# Add the data-points with noise in the X-axis
rug(jitter(X[,"x5"]))

```

```{r}
#Creating a density plot of output signal y
density_of_y <- density(Y)
hist(Y, freq = FALSE, main = "Histogram and density plot of y", xlab = "y Signal")
lines(density_of_y, lwd = 2, col = "red")
rug(jitter(Y))
```

### Task 1.3 Creating scatter plots to identify correlation arranging plot in single screen

```{r}
par(mfrow=c(2,2))

# Plotting input signal X1 against output signal Y
plot(X[,"x1"],Y,main = "Correlation betweeen X1 and Y signal", xlab = "X1 signal", ylab = "Output signal y")

# Plotting input signal X3 against output signal Y
plot(X[,"x3"],Y,main = "Correlation betweeen X3 and Y signal", xlab = "X3 signal", ylab = "Output signal y")

# Plotting input signal X4 against output signal Y
plot(X[,"x4"],Y,main = "Correlation betweeen X4 and Y signal", xlab = "X4 signal", ylab = "Output signal y")

# Plotting input signal X5 against output signal Y
plot(X[,"x5"],Y,main = "Correlation betweeen X5 and Y signal", xlab = "X5 signal", ylab = "Output signal y")
```

## Task 2

### calculating ones for binding the data

```{r}
ones = matrix(1 , length(X)/4,1)
head(ones)
```

## Task 2.1

### Calculating thetahat of each candidate model

```{r}
#Binding data from equation of model 1.
X_model1<-cbind(ones,X[,"x4"], X[,"x3"]^2)
head(X_model1)
```

```{r}
#Calculating thetahat of Model 1
Model1_thetahat=solve(t(X_model1) %*% X_model1) %*% t(X_model1) %*% Y
Model1_thetahat
```

```{r}
#Binding data from equation of model 2.
X_model2<-cbind(ones,X[,"x4"],X[,"x3"]^2,X[,"x5"])
head(X_model2)
```

```{r}
#Calculating thetahat of Model 2
Model2_thetahat=solve(t(X_model2) %*% X_model2) %*% t(X_model2) %*% Y
Model2_thetahat
```

```{r}
#Model 3
#Binding data from equation of model 3.
X_model3<-cbind(X[,"x3"],X[,"x4"],X[,"x5"]^3)
head(X_model3)
```

```{r}
#Calculating thetahat of Model 3
Model3_thetahat=solve(t(X_model3) %*% X_model3) %*% t(X_model3) %*% Y
Model3_thetahat
```

```{r}
#For model 4
#Binding data from equation of model 4.
X_model4<-cbind(ones,X[,"x4"],(X[,"x3"])^2,(X[,"x5"])^3)
head(X_model4)
```

```{r}
#Calculating thetahat of Model 4
Model4_thetahat=Ginv(t(X_model4) %*% X_model4) %*% t(X_model4) %*% Y
Model4_thetahat
```

```{r}
# for Model 5
#Binding data from equation of model 5.
X_model5<-cbind(ones,(X[,"x4"]),(X[,"x1"])^2,(X[,"x3"])^2)
head(X_model5)
```

```{r}
#Calculating thetahat of model 1
Model5_thetahat=solve(t(X_model5) %*% X_model5) %*% t(X_model5) %*% Y
Model5_thetahat
```

### Printing Value of theta of each model

```{r}
#model1
Model1_thetahat
```

```{r}
t(Model1_thetahat)
```

```{r}
#model 2
Model2_thetahat
```

```{r}
t(Model2_thetahat)
```

```{r}
#model 3
Model3_thetahat
```

```{r}
t(Model3_thetahat)
```

```{r}
#model 4
Model4_thetahat
```

```{r}
t(Model4_thetahat)
```

```{r}
#model 5
Model5_thetahat
```

```{r}
t(Model5_thetahat)
```
```{r}
output_dir <- "~/assignment/stat"

```

### Task 2.2

#### Calculating Y-hat and RSS for each model

```{r}
#Calculating Y-hat and RSS Model 1
Y_hat_model1 = X_model1 %*% Model1_thetahat
head(Y_hat_model1)
```

```{r}
#Calculating RSS
RSS_Model_1=sum((Y-Y_hat_model1)^2)
head(RSS_Model_1)
```

```{r}
# Calculating Y-hat and RSS of model 2
Y_hat_model2 = X_model2 %*% Model2_thetahat
head(Y_hat_model2)
```

```{r}
#Calculating RSS
RSS_Model_2=sum((Y-Y_hat_model2)^2)
head(RSS_Model_2)
```

```{r}
# Calculating Y-hat and RSS of model 3
Y_hat_model3 = X_model3 %*% Model3_thetahat
head(Y_hat_model3)
```

```{r}
#Calculating RSS
RSS_Model_3=sum((Y-Y_hat_model3)^2)
head(RSS_Model_3)
```

```{r}
# Calculating Y-hat and RSS of model 4
Y_hat_model4 = X_model4 %*% Model4_thetahat
head(Y_hat_model4)
```

```{r}
#Calculating RSS
RSS_Model_4=sum((Y-Y_hat_model4)^2)
head(RSS_Model_4)
```

```{r}
# Calculating Y-hat and RSS of model 5
Y_hat_model5 = X_model5 %*% Model5_thetahat
head(Y_hat_model5)
```

```{r}
#Calculating RSS
RSS_Model_5=sum((Y-Y_hat_model5)^2)
head(RSS_Model_5)
```

#### printing RSS value

```{r}
model1 <- c(RSS_Model_1)
model2 <- c(RSS_Model_2)
model3 <- c(RSS_Model_3)
model4 <- c(RSS_Model_4)
model5 <- c(RSS_Model_5)

dfRSS <- data.frame(model1, model2,model3,model4,model5)
dfRSS
```

#### Task 2.3 Calculating likelihood and variance of each model

```{r}
N=length(Y)

#Calculating the Variance of Model 1
Variance_model1=RSS_Model_1/(N-1)
Variance_model1
```

```{r}
#Calculating the log-likelihood of Model 1
likehood_Model_1=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model1))-(1/(2*Variance_model1))*RSS_Model_1
likehood_Model_1
```

```{r}
#Calculating Variance and log-likelihood of Model 2
Variance_model2=RSS_Model_2/(N-1)
Variance_model2
```

```{r}
likehood_Model_2=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model2))-(1/(2*Variance_model2))*RSS_Model_2
likehood_Model_2
```

```{r}
#Calculating Variance and log-likelihood of Model 3
Variance_model3=RSS_Model_3/(N-1)
Variance_model3
```

```{r}
likehood_Model_3=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model3))-(1/(2*Variance_model3))*RSS_Model_3
likehood_Model_3
```

```{r}
#Calculating Variance and log-likelihood of Model 4
Variance_model4=RSS_Model_4/(N-1)
Variance_model4
```

```{r}
likehood_Model_4=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model4))-(1/(2*Variance_model4))*RSS_Model_4
likehood_Model_4
```

```{r}
#Calculating Variance and log-likelihood of Model 5
Variance_model5=RSS_Model_5/(N-1)
Variance_model5
```

```{r}
likehood_Model_5=
  -(N/2)*(log(2*pi))-(N/2)*(log(Variance_model5))-(1/(2*Variance_model5))*RSS_Model_5
likehood_Model_5
```

#### Printing variance values

```{r}
model1 <- c(Variance_model1)
model2 <- c(Variance_model2)
model3 <- c(Variance_model3)
model4 <- c(Variance_model4)
model5 <- c(Variance_model5)

dfVariance <- data.frame(model1, model2,model3,model4,model5)
dfVariance
```

#### printing likelihood values

```{r}
model1 <- c(likehood_Model_1)
model2 <- c(likehood_Model_2)
model3 <- c(likehood_Model_3)
model4 <- c(likehood_Model_4)
model5 <- c(likehood_Model_5)

dfLikelihood <- data.frame(model1, model2,model3,model4,model5)
dfLikelihood
```

## Task 2.4

### Calculating AIC and BIC of each model

```{r}
# Calculating AIC and BIC of model 1
K_model1<-length(Model1_thetahat)
K_model1
```

```{r}
AIC_model1=2*K_model1-2*likehood_Model_1
AIC_model1
```

```{r}
BIC_model1=K_model1*log(N)-2*likehood_Model_1
BIC_model1
```

```{r}
## thetahat of model 2
K_model2<-length(Model2_thetahat)
K_model2
```

```{r}
##Calculating AIC and BIC of model 2
AIC_model2=2*K_model2-2*likehood_Model_2
AIC_model2
```

```{r}
BIC_model2=K_model2*log(N)-2*likehood_Model_2
BIC_model2
```

```{r}
## thetahat of model 3
K_model3<-length(Model3_thetahat)
K_model3
```

```{r}
##Calculating AIC and BIC of model 3
AIC_model3=2*K_model3-2*likehood_Model_3
AIC_model3
```

```{r}
BIC_model3=K_model3*log(N)-2*likehood_Model_3
BIC_model3
```

```{r}
## thetahat of model 4
K_model4<-length(Model4_thetahat)
K_model4
```

```{r}
##Calculating AIC and BIC of model 4
AIC_model4=2*K_model4-2*likehood_Model_4
AIC_model4
```

```{r}
BIC_model4=K_model4*log(N)-2*likehood_Model_4
BIC_model4
```

```{r}
## thetahat of model 5
K_model5<-length(Model5_thetahat)
K_model5
```

```{r}
##Calculating AIC and BIC of model 5
AIC_model5=2*K_model5-2*likehood_Model_5
AIC_model5
```

```{r}
BIC_model5=K_model5*log(N)-2*likehood_Model_5
BIC_model5
```

#### Printing K values

```{r}
model1 <- c(K_model1)
model2 <- c(K_model2)
model3 <- c(K_model3)
model4 <- c(K_model4)
model5 <- c(K_model5)

dfK <- data.frame(model1, model2,model3,model4,model5)
dfK
```

#### PRinting AIC Values

```{r}
model1 <- c(AIC_model1)
model2 <- c(AIC_model2)
model3 <- c(AIC_model3)
model4 <- c(AIC_model4)
model5 <- c(AIC_model5)

dfAIC <- data.frame(model1, model2,model3,model4,model5)
dfAIC
```

#### Printing BIC values

```{r}
model1 <- c(BIC_model1)
model2 <- c(BIC_model2)
model3 <- c(BIC_model3)
model4 <- c(BIC_model4)
model5 <- c(BIC_model5)

dfBIC <- data.frame(model1, model2,model3,model4,model5)
dfBIC
```

### Task 2.5 calculating error plotting normal/gussian distribution of each plot

```{r}
par(mfrow=c(1,1))

## Error of model1
model1_error <- Y-Y_hat_model1
head(model1_error)
```

```{r}
## Plotting the graph QQplot and QQ line of model 1
qqnorm(model1_error, col = "darkblue",main = "QQ plot of model 1")
qqline(model1_error, col = "red",lwd=1)
```

```{r}
## Error of model2
model2_error <- Y-Y_hat_model2 # error of model 2
## Plotting QQplot and QQ line of model 2
qqnorm(model2_error, col = "darkblue",main = "QQ plot of model 2")
qqline(model2_error, col = "red")
```

```{r}
## Error of model3
model3_error <- Y- Y_hat_model3
## Plotting QQplot and QQ line of model 3
qqnorm(model3_error, col = "darkblue",main = "QQ plot of model 3")
qqline(model3_error, col = "red")
```

```{r}
## Error of model4
model4_error <- Y-Y_hat_model4
## Plotting QQplot and QQ line of model 4
qqnorm(model4_error, col = "darkblue",main = "QQ plot of model 4")
qqline(model4_error, col = "red")
```

```{r}
## Error of model5
model5_error <- Y- Y_hat_model5
## Plotting QQplot and QQ line of model 5
qqnorm(model5_error, col = "darkblue",main = "QQ plot of model 5")
qqline(model5_error, col = "red")
```

### Task 2.7 Splitting data into training and testing dataset and calculating estimation based on training dataset

##### Also plotting normal distribution graph of training data

```{r}
## Spliting the dataset y into  Traning and testing data set.
split_Y<-initial_split(data = as.data.frame(Y),prop=.7)
## Traning splitted Y dataset 
Y_training_set<-training(split_Y)
Y_testing_set<-as.matrix(testing(split_Y))
## Testing splitted Y dataset 
Y_training_data<-as.matrix(Y_training_set)

## Spliting the dataset of X into  Traning and testing data set.
split_X<-initial_split(data = as.data.frame(X),prop=.7)
## Traning splitted X dataset
X_training_set<-training(split_X)
## Testing splitted X dataset 
X_testing_set<-as.matrix(testing(split_X))
X_testing_data<-as.matrix(X_testing_set)
X_training_data<-as.matrix(X_training_set)

### Estimating model parameters using Traning set
traning_ones=matrix(1 , length(X_training_set$x1),1)
# selected model 5 and using equation of model 5
X_traning_model<-cbind(traning_ones,X_training_set[,"x4"],(X_training_set[,"x1"])^2,(X_training_set[,"x3"])^2)
traning_thetahat=Ginv(t(X_traning_model) %*% X_traning_model) %*% t(X_traning_model) %*%  Y_training_data
  
### Model out/Prediction
Y_testing_hat = X_testing_data %*% traning_thetahat
head(Y_testing_hat)
```

```{r}
RSS_testing=sum((Y_testing_set-Y_testing_hat)^2)
head(RSS_testing)
```

```{r}
t.test(Y_training_data, mu=500, alternative="two.sided", conf.level=0.95)
```

```{r}
C_I1=454.0137
C_I2=454.8377
p2 <- plot(density(Y_training_data), col="blue", lwd=2,
           main="Distribution of Traning Data")
abline(v=C_I1,col="red", lty=2)
abline(v=C_I2,col="red", lty=2)
```

```{r}
thetaHat_training =solve(t(X_training_data) %*% X_training_data) %*% t(X_training_data) %*%Y_training_data
head(thetaHat_training)
```

```{r}
length(thetaHat_training)
```

```{r}
dis_test=density(Y_training_data)
plot((dis_test))
```

```{r}
plot(dis_test,main = "Density plot of Y Signal")
```

```{r}
### Calculating Confidential interval
z=1.96 ##(95%) Confidential interval
# error=((Y_testing_set-Y_testing_hat))
# n_len=length(Y_testing_hat)
# C_I_1= z * sqrt( (error * (1-error) ) / n_len)
# head(C_I_1)
# head(error)
# C_I_2= z * sqrt( (error * (1+error)) / n_len)
# head(C_I_2)        

Variance_model = mean((Y_testing_set - Y_testing_hat)^2)
se = sqrt(Variance_model)
CI_lower = Y_testing_hat - z * se
CI_upper = Y_testing_hat + z * se
head(CI_lower)
```

```{r}
head(CI_upper)
```

## Task 3

```{r}
## Model 5 will be used, parameter are selected and kept constant.
arr_1=0
arr_2=0
f_value=0
s_value=0
Model5_thetahat
```

```{r}
# Model 5 Parameters
theta0 <- 251.671653135 # selected parameter
theta1 <- 0.227168908   # selected parameter
theta2 <- -0.036521863  # fixed
theta3 <- -0.003650664  # fixed

# Simulation settings
epsilon <- RSS_Model_5 * 2
num_sim <- 10000

# Storage for accepted samples
accepted_theta0 <- c()
accepted_theta1 <- c()

# Rejection ABC
for (i in 1:num_sim) {
  sample_theta0 <- runif(1, theta0 * 0.8, theta0 * 1.2)
  sample_theta1 <- runif(1, theta1 * 0.8, theta1 * 1.2)
  theta_vector <- matrix(c(sample_theta0, sample_theta1, theta2, theta3))
  
  y_hat <- X_model5 %*% theta_vector
  rss_sim <- sum((Y - y_hat)^2)
  
  if (rss_sim < epsilon) {  # accept if close to observed
    accepted_theta0 <- c(accepted_theta0, sample_theta0)
    accepted_theta1 <- c(accepted_theta1, sample_theta1)
  }
}
par(mfrow=c(1,2))
# Plot Marginal Posterior Distributions
hist(accepted_theta0, breaks = 20, col = "orange", main = "Posterior of θ₀", xlab = "θ₀")
hist(accepted_theta1, breaks = 20, col = "lightblue", main = "Posterior of θ₁", xlab = "θ₁")

# Plot Joint Posterior Distribution
plot(
  accepted_theta0, accepted_theta1,
  main = "Joint Posterior Distribution",
  xlab = "θ₀", ylab = "θ₁",
  col = "darkblue", pch = 19
)

```

```{r}
mean(accepted_theta0); quantile(accepted_theta0, c(0.025, 0.975))
mean(accepted_theta1); quantile(accepted_theta1, c(0.025, 0.975))
```